---
title: "DS2 HW3"
author: "Jerry Zhou"
format: pdf
execute:
  message: false
  warning: false
  echo: true
editor: 
  markdown: 
    wrap: sentence
---

The `babynames` dataset in the `babynames` packages contains all births in the U.S. from 1880 to 2015 in which there were at least 5 births of that *sex-name* dyad in that year.
(A "dyad" is a combination of two variables, "triad" is three, etc.)

Create a dataset called `babynames_total` that totals `births` across all years.
(Note: The case in the new dataset should be `sex` and `name`, which means that each of these dyads should appear only once in the dataset.)

```{r}
library(tidyverse)
library(babynames)
head(babynames, 10)

babynames_total <- babynames |> 
  group_by(sex, name) |>
  summarise(births = sum(n), .groups = "drop") ; head(babynames_total, 10)

#.groups = "drop" prevents data from being grouped. Not an issue for graphs, but it is for manipulating data.
# also, why grouped by sex and not sex and name? 
```

Use this dataset to calculate what is asked below:

## 1) 10 most popular names

```{r}
babynames_total |>
  arrange(desc(births)) |>
  head(10)
```

## 2) Least popular names

a.  What is the smallest value of `births` in the dataset?
b.  How many *sex-name* have that many births?
c.  Based on the way the data were assembled, explain why there are so many.

```{r}
# a.
babynames_total |>
  arrange(births) |>
  head(10)

babynames_total |>
  pull(births) |>
  min()

# b.
babynames_total |> filter(births == 5)|> count()
babynames_total |> filter(births == 5)|> nrow()
```

-   5 births.
-   15,477 *sex-name* have 5 births.
-   The *sex-name* dyad contains all names and sexes recorded from 1880 to 2015. Even some of these rare names per sex can be repeated to have 5 births, especially if they're named that per generation (i.e. Jerry Sr. and Jerry Jr.).

ROB COMMENTS: 
I think that the arrange format for 2a definitely works, but as a way to simplify the process to find the smallest value, I just used the min() function inside a summarize() function to look something like (`summarize(min_ppl = min(births))`). This helps to simplify/reduce the number of lines of code!

## 3) 10 most popular [boys']{.underline} names that end in a vowel

```{r}
babynames_total_3 <- babynames_total |> mutate(last = str_sub(name,-1)); # babynames_total_3
babynames_total_3 |>
  filter(sex == "M") |>
  filter(str_detect(last, "[aeiou]")) |>
  arrange(desc(births)) |>
  head(10)
```

## 4) 10 most popular [girls']{.underline} names with joe/jo suffix

Find the 10 most popular girls' names that end in "joe" or "jo", excluding the names "Joe", and "Jo".
For this, problem, you should include all spellings, whether or not they are capitalized (so, for example, "Mary Jo" counts as ending in "jo")

```{r}
babynames_total_4 <- babynames_total |> mutate(last = str_sub(str_to_lower(name),-3))  #names start in lowercase
babynames_total_4 |>
  filter(sex == "F") |>
  filter(str_detect(last, "(jo|joe)$")) |> # $ only includes jo/joe names at end
  arrange(desc(births)) |>
  head(10)
```

ROB COMMENTS:
I really like the simplification of the code by using the `$` to indicate only jo/joe names at the end. I did it with slightly more lines of code and separate boolean expressions like `(last_two == "jo") | (last_three == "joe")` so I like the idea of doing it all at once.

## 5) Allnameslikethis

Did we need to worry about case in the last problem?
To answer this, verify the following:

a.  There are no spaces in any names
b.  The first letter is always capitalized
c.  No other letters are capitalized

```{r}
# a.
babynames_total_4 |>
  filter(str_detect(name, " ")) |>
  count()

# b. 
babynames_total_4 |>
  filter(str_sub(name, 1, 1) != str_to_upper(str_sub(name, 1, 1)) ) |>    #filter names starting with lowercase letters from names starting with uppercase letters
  count()

# c.
babynames_total_4 |>
  filter(str_detect(str_sub(name, 2), "[A-Z]")) |>  # starts with names on second letter
  count()

```

#### The problems below are optional, since they are not working with text, but I thought they provided an interesting challenge.

ROB COMMENTS:
I think that saving the above problem as an object helps expedite the process. I repeated my code from before and I'm now realizing how much more efficient it would be to just save the above manipulation as an object!

## 6) Closest to the average (optional)

Calculate the average number of births by *sex-name*.
Which *sex-name* is closest to that number?
Do this algorithmically in one pipeline, meaning no saving any objects and no typing in the average yourself.

```{r}
babynames_total |>
  mutate(avg_births = mean(births)) |>  # Calculates avg birth
  mutate(diff = abs(births - avg_births)) |>  # subtracts avg birth by birth in that
  arrange(diff) |> # arrange by lowest diff
  head(5)
```

* Newell (M) is the closest to the average number of births.

## 7) 10 most commmon cross-sex names (optional)

Show the most common names that are given both to males and to females.
You will have to make a decision about how to quantify this - I recommend not looking just at total births (male + female).

```{r}
babynames_total |>
  group_by(name) |>
  filter(sex == "F"| sex == "M") |>          # filter by both sexes
  summarise(
    male_births   = births[sex == "M"],
    female_births = births[sex == "F"],
    bothness = min(male_births, female_births),          # minimum name in either male or female births. Using max will heavily lean towards male names
    .groups = "drop"
  ) |>
  arrange(desc(bothness)) |>
  head(10)


# 45-55% male
babynames_total |>
  pivot_wider(names_from = sex, values_from = births) |>
  rename(female = F, male = M) |>
  filter(!is.na(female), !is.na(male)) |>
  mutate(prop_male = male / (male + female)) |>
  filter(0.45 < prop_male, prop_male < 0.55) |>
  arrange(desc(male+female))
```

ROB COMMENTS:
I think that the way you categorized the proportions of 45-55% male makes a lot of sense. It may be interesting if you explored a sensitivity analysis with a greater/lower proportion to see what the results would yield (perhaps 40-60%), but I think that the percentage you chose makes sense. Great job overall!